---
title: "AUtomated conservation assessment"
output: 
  html_document:
    theme: readable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
# Background
Species occurrence records can be used to approximate species ranges and generate preliminary conservation assessments. While comprehensive Red list assessments need a through case-by-case evaluation, preliminary assessments can help to speed up this process, by focussing on potentially threatened species.

# Objectives
After this exercise you will be abler to 
* Approximate species range sizes based on occurrence records
* Conduct an automated preliminary conservation assessment for multiple species based on occurrence records and Criterion B of the International Union for the conservation of Nature.

# Exercises
1. Calculate the area of the EOO for all species in your group (`CalcRangeSize`)
2. Do a preliminary conservation assessment of your group based on Criterion B using the ConR package. You can find a detailed tutorial [here](https://cran.r-project.org/web/packages/ConR/vignettes/my-vignette.html). (`IUCN.eval`)

# Library setup
```{r}
library(speciesgeocodeR)
library(ConR)
library(readr)
library(dplyr)
```

# Tutorial

## 1. Approximate species ranges

```{r}
dat <- read_csv("inst/occurrence_records_clean.csv")%>%
  dplyr::select(species, 
                decimallongitude = decimalLongitude,
                decimallatitude = decimalLatitude)

# Based on EOO
rs <- CalcRangeSize(dat)
```

A geospheric convex hull is a first approximation for a species range. However, some simple refinement might be desirable, for instance to limit the range only to biome where a given species has been recorded. This will take some time.

```{r}
# Limited to biomes with records
## Load Olson et al 2001 biomes
biom <- WWFload(x = "inst")
names(biom)

rs_biome <- CalcRangeSize(dat, biome = biom)

range <- data.frame(rs, rs_biome)
```

## 2. Automated conservation assessment 
You can use the ConR package for a preliminary conservation assessment orientated on the IUCN Red list Criterion B. T His is based on the EOO we have encountered above and additionally the Area of Occupancy and the number of subpopulations.

```{r, eval=FALSE}
# Format input data
inp <- dat%>%
  dplyr::select(ddlat = decimallatitude,
                ddlon = decimallongitude, 
                tax = species)


# Preliminary assessment
ev <- IUCN.eval(inp)
ev
```

## Write to disk

```{r}
write_csv(test, "inst/conservation_assessment.csv")
```





